The impact of information about tobacco-related reproductive vs. general health risks on South Indian women's tobacco use decisions

### 2020-11-06 paper.Rmd
Fixed the numbering of SI figures

---
title: The impact of information about tobacco-related reproductive vs. general health risks on South Indian women's tobacco use decisions
output: 
  bookdown::html_document2:
    toc: false
    toc_depth: 2
    number_sections: false
always_allow_html: true
csl: apa.csl
bibliography: mysore_tobacco2016.bib
---

@@

*Please address correspondence to Edward H. Hagen, Department of Anthropology, Washington State University, Vancouver, WA 98686. Phone number: 360-546-9257, email: edhagen@wsu.edu

@@

In our study, cotinine concentrations were bimodally distributed (see Figure \@ref(fig:cotdensity)). We therefore fit a Gaussian mixture model to log 10 cotinine values using the mixtools package [@benaglia2009]. Gaussian mixture models assume that each observation is generated by one of $K$ mixture components with probability $\pi_k$, where each component is a Gaussian distribution, $N(\mu_k, \sigma_k)$. The parameters $\{\mu_k, \sigma_k, \pi_k\}$ are estimated from the data using the iterative expectation-maximization (EM) algorithm [for details, see @benaglia2009]. In our data, using the default settings, the `normalmixEM` function determined that there were $K=2$ components, which we interpreted as tobacco non-users (left) vs. users (right) (see Figure \@ref(fig:cotdensity)).

```{r cotdensity, fig.cap="Histogram of log 10 cotinine concentrations overlaid with the two Gaussian components (red, green) estimated from the data using the `normalmixEM` function from mixtools."}
plot(m_mix, whichplots = 2, xlab2 = "Log10 cotinine concentration (ng/ml)")
```

We plotted the probability that each observation was generated by each component. The cotinine value for which there was an equal probability of belonging to the left vs. right component was `r cutoff` ng/ml (dotted line, Figure \@ref(fig:mixprob)), i.e., women with cotinine concentrations less than this were more likely to belong to the non-user component, whereas those with concentrations greater than this were more likely to belong to the user component.

```{r mixprob, fig.cap="Probabilities of belonging to the non-user (left) vs. user (right) component based on cotinine concentration. The dotted line represents the cotinine concentration for which there was an equal probability of belonging to either component."}
p_cot_mix
```

@@

We computed power using three sets of parameters for the linear function. First, we used the parameters estimated from our real (non-simulated) data (very similar to those reported in Table 2, Cotinine 2 model, except without a *Trimester* term). Important disadvantages of these parameters are that (1) they are noisy estimates of the true parameters, e.g., of the effect of the presentation on the slope (the interaction term), and hence there will be substantial uncertainty in the estimate of power, and (2) the estimated negative effect of the presentation on the slope was very likely biased toward a more negative value due to selection based on statistical significance. In addition, these parameters were surprising from a theoretical perspective. We predicted that the coefficient of baseline cotinine, $\beta_1$, would be about 1. Instead, it was about 2, i.e., the cotinine concentrations of participants in the GHP condition were about double their baseline levels. The effect of the RHP condition was to reduce this effect by about 75% (the coefficient of the interaction term, $\beta_3$, was about -1.5). The power to detect the (likely biased) effect of the RHP condition, i.e., the p-value for $\beta_3$, with our sample size of `r nrow(samples)/2`, and $\alpha=0.05$ was very high, `r round(100*pwr1['64'])`%. See Figure \@ref(fig:pwrcurves).

@@

```{r pwrcurves, fig.cap="Power versus sample size for three sets of parameters and $\\alpha=0.05$ (see text)."}
power_curves
```

@@

We used elastic net regression to fit a logistic regression model of tobacco use status as functions of our sociodemographic variables. Following standard procedure, we used 10-fold cross-validation to find the optimum value of $\lambda$, i.e., the one that minimized cross-validation error. We standardized our continuous variables (age, income, education years, number of children, domestic and non-domestic hours worked, and number of harms freelisted at baseline) by two standard deviations to approximate the standard deviations of our binary variables (married, pregnant, arranged marriage, and tobacco use by mother, mother-in-law, family, and friends), which improves comparisons of the coefficients of all variables [@gelman2008]. See Figure \@ref(fig:user).

```{r user, fig.cap=paste0("Logistic elasticnet regression coefficients of tobacco user status, fit using glmnet with $\\alpha=0.2$ and $\\lambda$ chosen by cross-validation. Continuous variables were centered and scaled by two standard deviations (see text). N=", nrow(dfw1))}
ggdotchart(exp(coef_user[-1,1])) + geom_vline(xintercept = 1, linetype = 'dotted') + scale_x_log10(breaks=seq(0.7, 1.7, by=0.2)) + xlab('\nOdds ratio')
```

@@

```{r cotharmfig, fig.width=10, fig.cap="Effects plot for a regression model of log10 cotinine concentration versus the change in number of (A) general harms and (B) reproductive harms mentioned at followup compared to baseline (see Table S1)."}
p_m_cot_deltargharmsGH + p_m_cot_deltargharmsRH + plot_annotation(tag_levels='A')
```

@@

```{r sharinglogistic, echo=F, message=F, warning=F, fig.cap="Logistic regression model of the effect of presentation type and the number of harms free-listed at followup on self-reported sharing of presentation with others. Base level is 'Did not share'. For regression coefficients, see Table 2, model Share presentation."}
plot(allEffects(m_share))
```

# Expanded table of regression models

>>>>

---
title: The impact of information about tobacco-related reproductive vs. general health risks on South Indian women's tobacco use decisions
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 2
    number_sections: false
always_allow_html: true
csl: apa.csl
bibliography: mysore_tobacco2016.bib
---

@@

*Address correspondence to Edward H. Hagen, Department of Anthropology, Washington State University, Vancouver, WA 98686. Phone number: 360-546-9257, email: edhagen@wsu.edu

@@

In our study, cotinine concentrations were bimodally distributed (see Figure S1). We therefore fit a Gaussian mixture model to log 10 cotinine values using the mixtools package [@benaglia2009]. Gaussian mixture models assume that each observation is generated by one of $K$ mixture components with probability $\pi_k$, where each component is a Gaussian distribution, $N(\mu_k, \sigma_k)$. The parameters $\{\mu_k, \sigma_k, \pi_k\}$ are estimated from the data using the iterative expectation-maximization (EM) algorithm [for details, see @benaglia2009]. In our data, using the default settings, the `normalmixEM` function determined that there were $K=2$ components, which we interpreted as tobacco non-users (left) vs. users (right) (see Figure S1).

```{r cotdensity}
plot(m_mix, whichplots = 2, xlab2 = "Log10 cotinine concentration (ng/ml)")
```

Figure S1: Histogram of log 10 cotinine concentrations overlaid with the two Gaussian components (red, green) estimated from the data using the `normalmixEM` function from mixtools.

We plotted the probability that each observation was generated by each component. The cotinine value for which there was an equal probability of belonging to the left vs. right component was `r cutoff` ng/ml (dotted line, Figure S2), i.e., women with cotinine concentrations less than this were more likely to belong to the non-user component, whereas those with concentrations greater than this were more likely to belong to the user component.

```{r mixprob}
p_cot_mix
```

Figure S2: Probabilities of belonging to the non-user (left) vs. user (right) component based on cotinine concentration. The dotted line represents the cotinine concentration for which there was an equal probability of belonging to either component.

@@

We computed power using three sets of parameters for the linear function. First, we used the parameters estimated from our real (non-simulated) data (very similar to those reported in Table 2, Cotinine 2 model, except without a *Trimester* term). Important disadvantages of these parameters are that (1) they are noisy estimates of the true parameters, e.g., of the effect of the presentation on the slope (the interaction term), and hence there will be substantial uncertainty in the estimate of power, and (2) the estimated negative effect of the presentation on the slope was very likely biased toward a more negative value due to selection based on statistical significance. In addition, these parameters were surprising from a theoretical perspective. We predicted that the coefficient of baseline cotinine, $\beta_1$, would be about 1. Instead, it was about 2, i.e., the cotinine concentrations of participants in the GHP condition were about double their baseline levels. The effect of the RHP condition was to reduce this effect by about 75% (the coefficient of the interaction term, $\beta_3$, was about -1.5). The power to detect the (likely biased) effect of the RHP condition, i.e., the p-value for $\beta_3$, with our sample size of `r nrow(samples)/2`, and $\alpha=0.05$ was very high, `r round(100*pwr1['64'])`%. See Figure S3.

@@

```{r pwrcurves}
power_curves
```

Figure S3: Power versus sample size for three sets of parameters and $\\alpha=0.05$ (see text).

@@

We used elastic net regression to fit a logistic regression model of tobacco use status as functions of our sociodemographic variables. Following standard procedure, we used 10-fold cross-validation to find the optimum value of $\lambda$, i.e., the one that minimized cross-validation error. We standardized our continuous variables (age, income, education years, number of children, domestic and non-domestic hours worked, and number of harms freelisted at baseline) by two standard deviations to approximate the standard deviations of our binary variables (married, pregnant, arranged marriage, and tobacco use by mother, mother-in-law, family, and friends), which improves comparisons of the coefficients of all variables [@gelman2008]. See Figure S4.

```{r user}
ggdotchart(exp(coef_user[-1,1])) + geom_vline(xintercept = 1, linetype = 'dotted') + scale_x_log10(breaks=seq(0.7, 1.7, by=0.2)) + xlab('\nOdds ratio')
```

Figure S4: `r paste0("Logistic elasticnet regression coefficients of tobacco user status, fit using glmnet with $\\alpha=0.2$ and $\\lambda$ chosen by cross-validation. Continuous variables were centered and scaled by two standard deviations (see text). N=", nrow(dfw1))`

@@

```{r cotharmfig}
p_m_cot_deltargharmsGH + p_m_cot_deltargharmsRH + plot_annotation(tag_levels='A')
```

Figure S5: Effects plot for a regression model of log10 cotinine concentration versus the change in number of (A) general harms and (B) reproductive harms mentioned at followup compared to baseline (see Table S1).

@@

```{r sharinglogistic, echo=F, message=F, warning=F}
plot(allEffects(m_share))
```

Figure S6: Logistic regression model of the effect of presentation type and the number of harms free-listed at followup on self-reported sharing of presentation with others. Base level is 'Did not share'. For regression coefficients, see Table 2, model Share presentation."

## Expanded table of regression models
